---
title: "GSS Network - Dyads and Triads"
format: gfm
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(paged.print = FALSE)
```

# Preparation

```{r}
# url1 <- "https://raw.githubusercontent.com/mahoffman/stanford_networks/main/data/comm59.dat.txt"

el <- read.table("../data/comm59.dat.txt", header = T)

# url2 <- "https://raw.githubusercontent.com/mahoffman/stanford_networks/main/data/comm59_att.dat.txt"

attributes <- read.table("../data/comm59_att.dat.txt", header = T)

# add an ID column
attributes$ID <- 1:nrow(attributes)
```

```{r}
#| message: false
library(igraph)
library(tidyverse)
```


```{r}
head(el)
```

```{r}
el_no_weight <- el %>%
  select(-rank) %>%
  mutate(across(where(is.numeric), as.character)) %>%
  as.matrix()

net59 <- graph_from_edgelist(el_no_weight,
  directed = T
)
```

```{r}
# First link vertex names to their place in the attribute dataset
linked_ids <- match(V(net59)$name, attributes$ID)

V(net59)$race <- attributes$race[linked_ids]
V(net59)$sex <- attributes$sex[linked_ids]
V(net59)$grade <- attributes$grade[linked_ids]
V(net59)$school <- attributes$school[linked_ids]

net59
```

# The Dyad

```{r}
edge_density(net59)
reciprocity(net59)
```

```{r}
random_graph <- sample_gnp(
  n = vcount(net59), p = edge_density(net59), directed = T
)
```

```{r random-graph-net59-1}
plot(random_graph,
  vertex.size = 2,
  vertex.label = NA,
  edge.curved = .1,
  vertex.color = "tomato",
  edge.arrow.size = .1,
  edge.width = .5,
  edge.color = "grey60"
)
```

```{r}
edge_density(net59)
reciprocity(net59)
edge_density(random_graph)
reciprocity(random_graph)
```

The reciprocity is much higher than expected.

```{r random-graph-net59-recip-1}
plot(net59,
  vertex.size = 2,
  vertex.label = NA,
  edge.curved = .1,
  vertex.color = "tomato",
  edge.arrow.size = .1,
  edge.width = .5,
  edge.color = "grey60"
)
```

# The Triad

```{r}
triad_census(net59)
triad_census(random_graph)
```

Triads are common in the observed network but not in the random graph.

# Distribution of random graphs

```{r}
trial <- vector("list", 100)

for (i in 1:length(trial)) {
  trial[[i]] <- sample_gnp(
    n = vcount(net59), p = edge_density(net59), directed = T
  ) %>%
    triad_census()
}

trial_df <- do.call("rbind", trial)

colnames(trial_df) <- c("003", "012", "102", "021D", "021U", "021C", "111D", "111U", "030T", "030C", "201", "120D", "120U", "120C", "210", "300")


trial_df_w_observed <- rbind(trial_df, as.numeric(triad_census(net59)))
```

```{r}
head(trial_df_w_observed)
```

Let’s produce, for each column, some simple statistics, like a mean and a confidence interval. First, standardize all of the columns by dividing each of their values by the largest value in that column, so that each will be on a similar scale (0 to 1), we can visualize them meaningfully 

```{r}
trial_df_w_observed <- as.data.frame(trial_df_w_observed)

trial_df_w_observed[, 1:ncol(trial_df_w_observed)] <- sapply(
  trial_df_w_observed[, 1:length(trial_df_w_observed)],
  function(x) x / max(x)
)

trial_df <- as.data.frame(trial_df_w_observed[1:100, ])
observed <- as.numeric(trial_df_w_observed[101, ])
```

```{r}
summarized_stats <- data.frame(
  TriadType = colnames(trial_df),
  Means = sapply(trial_df, mean),
  LowerCI = sapply(trial_df, function(x) quantile(x, 0.05)),
  UpperCI = sapply(trial_df, function(x) quantile(x, 0.95)),
  Observed = observed
)
summarized_stats
```

```{r mna-dyadtriad-plot-1}
library(ggplot2)
ggplot(summarized_stats) +
  geom_path(aes(x = TriadType, y = Means, color = TriadType)) +
  geom_errorbar(aes(
    x = TriadType, ymin = LowerCI, ymax = UpperCI,
    color = TriadType
  ), width = .1) +
  geom_point(aes(x = TriadType, y = Observed, color = "Observed")) +
  coord_flip()
```

# Banning Triads

```{r}
# A basic function which prevents the formation of specified triads in a random graph simulation
banning_triads_game <- function(n = 100, porm = .05,
                                banned = c(2), sim_max = 1000000,
                                probrecip = .5) {
  if (any(c(1) %in% banned)) {
    stop("Can't ban 003s") # Stops the simulation if the user tried to ban 003 triads
  }

  num_edges <- round(n * (n - 1) * porm, 2) # calculates the desired number of edges according to the N and Porm parameters

  net <- make_empty_graph(n = n, directed = TRUE) # initializes an empty network

  edge_count <- 0
  sim_count <- 0

  while (edge_count < num_edges) {
    # This part samples two nodes, checks whether the two sampled nodes are the same node, and whether an edge is already present in the network between these nodes

    uniq <- TRUE
    edge_present <- TRUE
    while (uniq == TRUE | edge_present == TRUE) {
      edge_id <- sample(1:n, 2, replace = T)
      uniq <- edge_id[1] == edge_id[2]
      reciprocated <- sample(c(FALSE, TRUE), 1, prob = c(1 - probrecip, probrecip))
      edge_present_1 <- are_adjacent(net, edge_id[1], edge_id[2])
      if (reciprocated) {
        edge_present_2 <- are_adjacent(net, edge_id[2], edge_id[1])
        edge_present <- edge_present_1 | edge_present_2
      } else {
        edge_present <- edge_present_1
      }
    }

    # Calculates the traid census for the network before adding an edge
    before <- triad.census(net)
    net_new <- net + edge(edge_id) # Adds in the edge
    if (reciprocated) {
      edge_id_rev <- edge_id[2:1]
      net_new <- net_new + edge(edge_id_rev) # Adds in the edge
    }
    after <- triad.census(net_new) # Calculates the triad census again
    triad_diff <- after - before # Checks to see how much the triad census changed

    if (all(triad_diff[banned] == 0)) {
      net <- net_new # If the banned triads still aren't observed, then the new network is accepted.
    }

    edge_count <- ecount(net) # number of edges updated
    sim_count <- sim_count + 1 # Simulation count updated
    if (sim_count > sim_max) {
      print("Warning: Failed to converge, banned triads may be incompatible") # exits simulation if simulation max count is exceeded
      return(net)
    }
  }
  return(net) # Returns the simulated network
}
```

```{r}
no_cycles <- banning_triads_game(banned = c(4, 5, 7))
triad_census(no_cycles)
```

```{r mna-dyadtriad-plot-2}
plot(no_cycles,
  vertex.size = 2, vertex.label = NA,
  vertex.color = "tomato", edge.arrow.size = .2
)
```
